---
title: "Lab 5"
output: html_document
date: "`r Sys.Date()`"
---

## Overview

This lab follows a simple workflow:

1. use a linear regression to describe the conditional expectation function,
2. interpret selected coefficients as marginal effects,
3. show how omitted-variable bias changes the coefficient of interest,
4. use sensitivity analysis to ask how strong an unobserved confounder would need to be to change the conclusion,
5. briefly illustrate how the `regweight` package can be used to inspect the implicit regression weights behind a coefficient.

We use the LaLonde job-training data from the `CBPS` package throughout.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r packages}
install.packages("remotes")

remotes::install_github(
  "ddimmery/regweight",
  upgrade = "never",
  dependencies = TRUE,
  build_vignettes = FALSE
)
packages <- c("dplyr", "ggplot2", "broom", "knitr", "CBPS", "sensemakr", "regweight")
missing_pkgs <- packages[!vapply(packages, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_pkgs) > 0) {
  stop(
    "Please install these packages before knitting: ",
    paste(missing_pkgs, collapse = ", ")
  )
}

invisible(lapply(packages, library, character.only = TRUE))
```

## 1. Data preparation

The outcome is earnings in 1978. The treatment is participation in the training program. We also include a standard set of pre-treatment controls.

```{r data-prep}
data("LaLonde", package = "CBPS")

df <- LaLonde %>%
  filter((exper == 1 & treat == 1) | (exper == 0 & treat == 0)) %>%
  mutate(
    log_re78 = log(re78 + 1),
    re74_log = log(re74 + 1),
    re75_log = log(re75 + 1),
    treat = as.numeric(treat)
  )

analysis_vars <- c(
  "log_re78", "treat", "age", "educ", "black", "hisp",
  "married", "nodegr", "re74_log", "re75_log", "re74.miss"
)

summary_df <- df %>%
  dplyr::select(dplyr::all_of(analysis_vars)) %>%
  summary()

summary_df
```

## 2. Conditional expectation function and marginal effects

Here, `treat` is binary, so the coefficient on `treat` is interpreted as the conditional difference in expected log earnings between treated and untreated units, holding the observed covariates fixed.

```{r main-model}
main_formula <- log_re78 ~ treat + age + educ + black + hisp + married +
  nodegr + re74_log + re75_log + re74.miss

m_full <- lm(main_formula, data = df)

broom::tidy(m_full) %>%
  knitr::kable(digits = 3, caption = "Main regression")
```

### Interpreting marginal effects

For a binary treatment, the marginal effect is a conditional mean difference. For a continuous regressor in a linear model, the coefficient itself is the marginal effect.

```{r marginal-effects}
treat_effect <- coef(m_full)["treat"]
age_effect <- coef(m_full)["age"]

marginal_effects <- tibble::tibble(
  quantity = c("Effect of treatment (binary D)", "Marginal effect of age (continuous X)"),
  estimate = c(treat_effect, age_effect)
)

marginal_effects %>%
  knitr::kable(digits = 3, caption = "Two coefficient interpretations in the linear model")
```

We can also show the treatment effect using predictions for the same covariate profile under `treat = 0` and `treat = 1`.
In this example, our profile is the average values of the covariates in the sample. 
The difference in predicted log earnings between `treat = 1` and `treat = 0` is the same as the coefficient on `treat` in the linear model, which is a special case of the linearity assumption. 
In a more flexible model, these two quantities would not be identical, but they would still be related to each other as marginal effects and predictions are in general.

```{r prediction-example}
profile <- df %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    educ = mean(educ, na.rm = TRUE),
    black = mean(black, na.rm = TRUE),
    hisp = mean(hisp, na.rm = TRUE),
    married = mean(married, na.rm = TRUE),
    nodegr = mean(nodegr, na.rm = TRUE),
    re74_log = mean(re74_log, na.rm = TRUE),
    re75_log = mean(re75_log, na.rm = TRUE),
    re74.miss = mean(re74.miss, na.rm = TRUE)
  )

pred_data <- bind_rows(
  mutate(profile, treat = 0),
  mutate(profile, treat = 1)
)

pred_data$pred_log_re78 <- predict(m_full, newdata = pred_data)
pred_data

pred_diff <- diff(pred_data$pred_log_re78)
pred_diff
```

## 3. Omitted-variable bias

We now compare the treatment coefficient across models with different control sets.

- `m_naive`: no controls.
- `m_demo`: demographics only.
- `m_short`: adds 1974 earnings, but still omits 1975 earnings.
- `m_full`: adds 1975 earnings as well.

```{r ovb-models}
m_naive <- lm(log_re78 ~ treat, data = df)

m_demo <- lm(
  log_re78 ~ treat + age + educ + black + hisp + married + nodegr,
  data = df
)

m_short <- lm(
  log_re78 ~ treat + age + educ + black + hisp + married + nodegr +
    re74_log + re74.miss,
  data = df
)

model_compare <- bind_rows(
  broom::tidy(m_naive) %>% filter(term == "treat") %>% mutate(model = "Naive"),
  broom::tidy(m_demo) %>% filter(term == "treat") %>% mutate(model = "Demographics only"),
  broom::tidy(m_short) %>% filter(term == "treat") %>% mutate(model = "Omit re75_log"),
  broom::tidy(m_full) %>% filter(term == "treat") %>% mutate(model = "Full model")
) %>%
   dplyr::select(model, estimate, std.error, statistic, p.value)

model_compare %>%
  knitr::kable(digits = 3, caption = "How the treatment coefficient changes across specifications")
```

### Exact decomposition for one omitted variable

Now we use the omitted-variable bias identity with one omitted covariate: `re75_log`.

- `delta_tilde`: effect of `re75_log` on the outcome in the long regression.
- `gamma_tilde`: residual association between `re75_log` and `treat`, conditional on the other controls.
- `delta_tilde * gamma_tilde`: implied bias from omitting `re75_log`.

```{r ovb-decomposition}
controls_z <- c("age", "educ", "black", "hisp", "married", "nodegr", "re74_log", "re74.miss")

aux_formula <- as.formula(
  paste("re75_log ~ treat +", paste(controls_z, collapse = " + "))
)

aux_model <- lm(aux_formula, data = df)

delta_tilde <- coef(m_full)["re75_log"]
gamma_tilde <- coef(aux_model)["treat"]

bias_from_formula <- delta_tilde * gamma_tilde
bias_from_models <- coef(m_short)["treat"] - coef(m_full)["treat"]

ovb_table <- tibble::tibble(
  quantity = c(
    "delta_tilde: effect of re75_log on log_re78",
    "gamma_tilde: residual imbalance of re75_log with treat",
    "delta_tilde * gamma_tilde",
    "Difference between short and full treatment coefficients"
  ),
  value = c(delta_tilde, gamma_tilde, bias_from_formula, bias_from_models)
)

ovb_table %>%
  knitr::kable(digits = 4, caption = "OVB decomposition for omitting re75_log")
```

The last two rows should be almost identical up to numerical precision. That is the linear omitted-variable-bias decomposition at work.

## 4. Partial (R^2) and sensitivity analysis

Sensitivity analysis asks how strong an unobserved confounder would need to be to change the treatment estimate. The two key quantities are:

- outcome-side partial (R^2): how much of the residual outcome variation a confounder explains,
- treatment-side partial (R^2): how much of the residual treatment variation it explains.

We first compute the partial (R^2) of an observed benchmark variable, `re75_log`, and then use those values to calibrate hypothetical confounders.

### A helper for partial (R^2)

```{r partial-r2-helper}
partial_r2 <- function(reduced_model, full_model) {
  r2_reduced <- summary(reduced_model)$r.squared
  r2_full <- summary(full_model)$r.squared
  (r2_full - r2_reduced) / (1 - r2_reduced)
}
```

### Benchmark partial (R^2) for `re75_log`

```{r benchmark-r2}
# Outcome-side partial R^2 for re75_log
m_y_reduced <- lm(
  log_re78 ~ treat + age + educ + black + hisp + married + nodegr +
    re74_log + re74.miss,
  data = df
)

m_y_full <- lm(
  log_re78 ~ treat + age + educ + black + hisp + married + nodegr +
    re74_log + re74.miss + re75_log,
  data = df
)

r2_y_benchmark <- partial_r2(m_y_reduced, m_y_full)

# Treatment-side partial R^2 for re75_log
m_d_reduced <- lm(
  treat ~ age + educ + black + hisp + married + nodegr + re74_log + re74.miss,
  data = df
)

m_d_full <- lm(
  treat ~ age + educ + black + hisp + married + nodegr + re74_log + re74.miss + re75_log,
  data = df
)

r2_d_benchmark <- partial_r2(m_d_reduced, m_d_full)

benchmark_table <- tibble::tibble(
  benchmark = "re75_log",
  outcome_side_partial_r2 = r2_y_benchmark,
  treatment_side_partial_r2 = r2_d_benchmark
)

benchmark_table %>%
  knitr::kable(digits = 4, caption = "Benchmark partial R-squared values")
```

### Manual contour plot

The graph below is built from a grid of hypothetical confounders. 
Each point in the graph is a pair of partial (R^2) values for the treatment and outcome regressions, respectively.
For each pair we compute the implied bias using the formula from Cinelli and Hazlett (2020).
The red dashed contour is the set of confounder strengths that would reduce the treatment estimate to zero.

```{r sensitivity-grid}
estimate_treat <- coef(m_full)["treat"]
se_treat <- summary(m_full)$coefficients["treat", "Std. Error"]
df_treat <- df.residual(m_full)

sens_grid <- expand.grid(
  r2_treatment = seq(0.001, 0.25, length.out = 150),
  r2_outcome = seq(0.001, 0.25, length.out = 150)
) %>%
  mutate(
    implied_bias = se_treat * sqrt((r2_outcome * r2_treatment / (1 - r2_treatment)) * df_treat),
    adjusted_estimate = estimate_treat - implied_bias
  )

# Simple didactic benchmark scaling for plotting.
# This is useful for teaching the logic of the graph.
benchmark_points <- tibble::tibble(
  label = c("1x re75_log", "2x re75_log"),
  r2_treatment = pmin(c(1, 2) * r2_d_benchmark, 0.99),
  r2_outcome = pmin(c(1, 2) * r2_y_benchmark, 0.99)
) %>%
  mutate(
    implied_bias = se_treat * sqrt((r2_outcome * r2_treatment / (1 - r2_treatment)) * df_treat),
    adjusted_estimate = estimate_treat - implied_bias
  )

benchmark_points %>%
  knitr::kable(digits = 4, caption = "Benchmark-calibrated hypothetical confounders")
```

```{r sensitivity-plot, fig.width=8, fig.height=6}
ggplot(sens_grid, aes(x = r2_treatment, y = r2_outcome, z = adjusted_estimate)) +
  geom_contour(color = "grey55", breaks = pretty(sens_grid$adjusted_estimate, n = 6)) +
  geom_contour(color = "red", linetype = "dashed", linewidth = 1, breaks = 0) +
  geom_point(
    data = benchmark_points,
    aes(x = r2_treatment, y = r2_outcome),
    inherit.aes = FALSE,
    size = 2
  ) +
  geom_text(
    data = benchmark_points,
    aes(x = r2_treatment, y = r2_outcome, label = paste0(label, "\nAdj. est = ", round(adjusted_estimate, 3))),
    inherit.aes = FALSE,
    nudge_y = 0.012,
    size = 3.2
  ) +
  labs(
    title = "Sensitivity-analysis contour plot",
    x = expression(R[D %~% U %.% X]^2),
    y = expression(R[Y %~% U %.% D + X]^2),
    caption = "Grey curves: constant adjusted estimates. Red dashed curve: combinations that reduce the treatment estimate to zero."
  ) +
  theme_minimal()
```

### `sensemakr` implementation

The manual contour plot above is useful for understanding the logic. The `sensemakr` package automates a formal version of this exercise.

```{r sensemakr-example}
sens <- sensemakr::sensemakr(
  model = m_full,
  treatment = "treat",
  benchmark_covariates = "re75_log",
  kd = c(1, 2)
)

sens
```

```{r sensemakr-plot, fig.width=8, fig.height=6}
plot(sens)
```

## 5. Brief introduction to `regweight`

The `regweight` package is useful when you want to know **which observations matter most for a particular regression coefficient**. 
Instead of focusing on global leverage, it focuses on the implicit weights behind one selected term.

Below we ask for the implicit regression weights behind the coefficient on `treat` in the full model.

```{r regweight-basic}
rw_treat <- regweight::calculate_weights(m_full, "treat")

hist(rw_treat)
```
The histogram above shows the distribution of the implicit regression weights for the `treat` coefficient.
Some observations have more influence on the `treat` coefficient than others, and this is not necessarily related to leverage in the usual sense.

We can also inspect how those implicit weights vary over a discrete covariate.
```{r regweight-discrete}
plot(rw_treat, df$married)
```
The implicit sample for the `treat` coefficient gives more weight to some observations than others, and this can be related to covariate values.
The nominal sample is the original data, and the implicit sample is the data weighted by the regression weights for the `treat` coefficient.
In our example, the implicit sample for the `treat` coefficient gives more weight to unmarried units than married units. This means that the `treat` coefficient is learning more from unmarried units than married units.

And we can compare the nominal and implicit samples for a few covariates.

```{r regweight-summary}
summary(
  rw_treat,
  dplyr::select(df, age, married, nodegr, re74_log)
)

plot(rw_treat, df$age)
```
For the `age` variable, the implicit sample is younger than the nominal sample, which means that the `treat` coefficient is learning more from younger units than older units.
From the plot of `age`, we can see that the implicit sample has more weight on younger units than older units, which is consistent with the summary statistics.

### Interpretation

A useful reading strategy is:

- without using weights, we can only talk about the regression coefficient as a global quantity that describes the entire sample,
- with regression weights, we can talk about the regression coefficient as a local quantity that describes a particular implicit sample,
- if the weight distribution is very dispersed, the regression coefficient is learning from a wide range of observations,
- if the weight distribution is very concentrated, a relatively small set of observations is driving the coefficient,
- if the implicit sample looks very different from the nominal sample, the regression coefficient may be learning mostly from a specific part of the data,
- this is especially useful when you worry that the OLS estimand may be giving more influence to some units than others.


