---
title: "Causal graphs with dagitty in R"
author: "Draft template"
date: "2026-02-13"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Core packages for DAGs
# install.packages(c("dagitty", "ggdag", "tidyverse"))
library(dagitty)
library(ggdag)
library(tidyverse)
```

## Goal and roadmap

This R Markdown walks through a **step-by-step workflow** to build and interpret causal graphs in R:

1. Start with the **three canonical motifs**:
   - *Mediator* (chain)
   - *Confounder* (fork / common cause)
   - *Collider* (common effect)

2. For each motif:
   - draw the DAG
   - check the implied (conditional) independencies
   - connect it to a **real-world example**

3. Recreate two **more complex** examples, then connect them to a **real applied example**.

# Part 1 — Three canonical motifs

## 1) Mediator (chain):  X → Z → Y

### DAG

```{r}
dag_med <- dagitty("dag {
  X -> Z -> Y
}")

ggdag(dag_med) + theme_dag()
```

```{r}
# dagify version for better styling
dag_med <- dagify(
  Z ~ X,
  Y ~ Z,
  exposure = "X",
  outcome  = "Y",
  labels = c(X = "X", Y = "Y", Z = "Z")
)
# Add styling for better visualization
ggdag(dag_med, text = TRUE) +
  geom_dag_edges(edge_width = 1.3) +
  geom_dag_point(size = 11) +
  geom_dag_text(color = "white", size = 5) +
  theme_dag() +
  theme(
    plot.margin = margin(20, 20, 20, 20),
    plot.title = element_text(face = "bold")
  ) +
  ggtitle("Mediator: conditioning on Z blocks the path from X to Y")
```
### Implied independencies and conditioning

- In a simple chain, **X and Y are associated** (because X can reach Y through Z).
- But **conditioning on Z blocks** the directed ("front-door") path between X and Y.

```{r}
impliedConditionalIndependencies(dag_med)

# Check d-separation explicitly:
dseparated(dag_med, "X", "Y")                 # FALSE (dependent)
dseparated(dag_med, "X", "Y", "Z")            # TRUE  (independent given Z)
```

### Real-world example (mediator)

**Policy training program (X)** increases **skills (Z)**, which increases **wages (Y)**.

- If you *control for skills*, you are estimating something closer to the **direct effect** of training on wages, not the **total effect**.

---

## 2) Confounder (fork):  X ← Z → Y

### DAG

```{r}
dag_confound <- dagitty("dag {
  Z -> X
  Z -> Y
}")

ggdag(dag_confound) + theme_dag()
```

```{r}
# dagify version for better styling
dag_confound <- dagify(
  X ~ Z,
  Y ~ Z,
  exposure = "X",
  outcome  = "Y",
  labels = c(X = "X", Y = "Y", Z = "Z")
)

# Add styling for better visualization
ggdag(dag_confound, text = TRUE) +
  geom_dag_edges(edge_width = 1.3) +
  geom_dag_point(size = 11) +
  geom_dag_text(color = "white", size = 5) +
  theme_dag() +
  theme(
    plot.margin = margin(20, 20, 20, 20),
    plot.title = element_text(face = "bold")
  ) +
  ggtitle("Confounder: conditioning on Z blocks the backdoor path between X and Y")
```

### Implied independencies and conditioning

- Without conditioning, **X and Y are associated** through their common cause Z.
- Conditioning on Z blocks the backdoor path.

```{r}
impliedConditionalIndependencies(dag_confound)

dseparated(dag_confound, "X", "Y")         # FALSE (dependent)
dseparated(dag_confound, "X", "Y", "Z")    # TRUE  (independent given Z)

ggdag_adjustment_set(dag_confound, node_size = 20) +
  theme_dag()
```

### Real-world example (confounder)

**Coffee (X)** and **heart disease (Y)** may look associated because **smoking (Z)** increases coffee consumption and heart disease.

- Adjusting for smoking targets the causal link from coffee to heart disease (if no other backdoors remain).

---

## 3) Collider (common effect):  X → Z ← Y

### DAG

```{r}
dag_collider <- dagitty("dag {
  X -> Z
  Y -> Z
}")

ggdag(dag_collider) + theme_dag()
```

```{r}
# dagify version for better styling
dag_collider <- dagify(
  Z ~ X + Y,
  exposure = "X",
  outcome  = "Y",
  labels = c(X = "X", Y = "Y", Z = "Z")
)

# Add styling for better visualization
ggdag(dag_collider, text = TRUE) +
  geom_dag_edges(edge_width = 1.3) +
  geom_dag_point(size = 11) +
  geom_dag_text(color = "white", size = 5) +
  theme_dag() +
  theme(
    plot.margin = margin(20, 20, 20, 20),
    plot.title = element_text(face = "bold")
  ) +
  ggtitle("Collider: conditioning on Z opens association between X and Y")
```

### Implied independencies and conditioning

- In a pure collider structure, **X and Y are independent marginally**.
- Conditioning on the collider **opens** a path and makes them dependent (*collider bias* / Berkson’s paradox).

```{r}
impliedConditionalIndependencies(dag_collider)

dseparated(dag_collider, "X", "Y")          # TRUE  (independent)
dseparated(dag_collider, "X", "Y", "Z")     # FALSE (dependent given Z)
```

### Real-world example (collider)

**Talent (X)** and **connections (Y)** both increase the probability of being **hired (Z)**.

- If you only analyze people **who were hired** (conditioning on Z), talent and connections can become spuriously related.

---

# Part 2 — Turning motifs into “causal adjustment” questions

A key workflow with DAGs is:

1. Encode your causal story in a DAG.
2. Choose the causal estimand: total effect? direct effect? etc.
3. Use `adjustmentSets()` (backdoor) to propose controls.
4. Think hard: are those controls measured well? are they pre-treatment?

## Example: asking dagitty for adjustment sets

```{r}
# In a chain, Z is not a valid adjustment set for the total effect of X on Y (it blocks the path).
adjustmentSets(dag_med, exposure = "X", outcome = "Y")

# In a fork, Z is a valid adjustment set for the effect of X on Y.
adjustmentSets(dag_confound, exposure = "X", outcome = "Y")

# In a collider, there are no valid adjustment sets for the effect of X on Y (conditioning on Z creates bias).
adjustmentSets(dag_collider, exposure = "X", outcome = "Y")
```

---

# Part 3 — Two more complex examples (inspired by the blackboard)

## 4) Mediation + unobserved confounding (U)

Here is a common “realistic” picture:

- Treatment **D** affects mediator **M**, which affects outcome **Y**
- There is an unobserved confounder **U** that affects both **D** and **Y**

```{r}
dag_complex1 <- dagitty("dag {
  D -> M -> Y
  U -> D
  U -> Y
}")

ggdag(dag_complex1) + theme_dag()
```

### What does the DAG imply?

```{r}
impliedConditionalIndependencies(dag_complex1)
adjustmentSets(dag_complex1, exposure = "D", outcome = "Y")
```

**Interpretation**

- The backdoor path **D ← U → Y** is open.
- Because **U is unobserved**, there is **no valid adjustment set using observed variables only**.
- This is when you start thinking about:
  - instruments,
  - natural experiments,
  - panel designs / fixed effects with strong assumptions,
  - negative controls / sensitivity analysis,
  - or (sometimes) collecting better covariates.

> Note: controlling for the mediator **M** here *does not* solve confounding; it changes the estimand (direct effect) and may create additional bias depending on what else is in the system.

---

## 5) Why adjusting for a confounder does not make D and Y independent

A frequent confusion:

- Under ignorability / unconfoundedness, we want something like  
  **D ⟂ (Y(0), Y(1)) | U**  
  (treatment is independent of **potential outcomes**, given confounders)

That does **not** mean  
**D ⟂ Y | U**,  
because **Y depends on D** through the causal effect.

We can show this with a DAG:

```{r}
# Causal effect: D -> M -> Y
# Confounder: U1 affects D and Y
# Upstream driver: U2 affects U1 and Y

dag_complex2 <- dagitty("dag {
  D -> M -> Y
  U1 -> D
  U1 -> Y
  U2 -> U1
  U2 -> Y
}")


ggdag(dag_complex2) + theme_dag()

# dagify version for better styling
dag_complex2 <- dagify(
  M ~ D,
  Y ~ M + U1 + U2,
  U1 ~ U2,
  D ~ U1,
  exposure = "D",
  outcome  = "Y",
  labels = c(D = "D", M = "M", Y = "Y", U1 = "U1", U2 = "U2")
)

ggdag_adjustment_set(dag_complex2, node_size = 20) +
  theme_dag()
```

### Key takeaway via d-separation

Even conditioning on U1, **D is not d-separated from Y** because the directed causal path remains.

```{r}
impliedConditionalIndependencies(dag_complex2)
dseparated(dag_complex2, "D", "Y", "U1")          # FALSE
dseparated(dag_complex2, "D", "Y", c("U1","M"))   # TRUE (but note: conditioning on M changes estimand. To get the total effect, we would not condition on M.)
adjustmentSets(dag_complex2, exposure = "D", outcome = "Y")  # U1 is a valid adjustment set for the total effect, but it does not make D and Y independent because of the causal path through M.
```

**Interpretation**

- Adjusting for confounders removes *spurious* association (backdoor paths),
  but the *causal* association (frontdoor path) remains.

---

# Part 4 — A worked real example (template you can adapt)

Below is one realistic applied story (feel free to swap it for your own application).

## Example: land-tenure formalization and deforestation

- **D** = land-tenure formalization intervention (e.g., titling / cadaster update)
- **M** = perceived tenure security / investment (mechanism)
- **Y** = deforestation rate
- **U1** = baseline accessibility / road proximity (measured)
- **U2** = illicit armed group pressure / illegal market dynamics (often imperfectly measured)

```{r}
dag_defor <- dagify(
  M ~ D,
  Y ~ M + U1 + U2,
  D ~ U1 + U2,
  exposure = "D",
  outcome  = "Y",
  labels = c(D = "Land-tenure formalization (D)", M = "Perceived tenure security (M)", Y = "Deforestation rate (Y)", U1 = "Baseline accessibility (U1)", U2 = "Illicit market dynamics (U2)")
)

ggdag(dag_defor) + theme_dag()

ggdag_adjustment_set(dag_defor, node_size = 20) +
  theme_dag()
```

### Identification question

```{r}
adjustmentSets(dag_defor, exposure = "D", outcome = "Y")
```

- If both **U1 and U2** were measured well, `{U1, U2}` would block backdoors.
- In practice, **U2** may be partially unobserved → motivates designs beyond plain regression adjustment.

---

# Appendix — Useful functions / snippets

## Quick checklist for any DAG

```{r}
# 1) Write the DAG
# dag <- dagitty("dag { ... }")

# 2) dagify version for better styling
# dag <- dagify(
#   ...,
#   exposure = "X", # optional but helps with ggdag styling
#   outcome  = "Y",
#   labels = c(X = "X", Y = "Y", Z = "Z") # optional but helps with ggdag styling
# )

# 3) Plot
# ggdag(dag, text = FALSE) + theme_dag()

# 4) See implied independencies
# impliedConditionalIndependencies(dag)

# 5) Find minimal backdoor adjustment sets
# adjustmentSets(dag, exposure = "X", outcome = "Y")

# 6) Visualize adjustment sets
# ggdag_adjustment_set(dag, node_size = 20) + theme_dag()

# 7) Test specific d-separation claims
# dseparated(dag, "X", "Y", "Z")  # or conditioning set as a vector
```

