---
title: "Causal Inference Lab 6: Weighting, and Balance"
output: html_document
fig_caption: yes
editor_options:
  chunk_output_type: console
---

## Overview

In this lab, we move from the theory of weighting to a practical application using the LaLonde-Dehejia-Wahba (LDW) data. The main goal is to see how different estimators behave when treated and control units are not directly comparable.

We will focus on four ideas:

1. **Balance** as the design target.
2. **Propensity scores** as a balancing score and overlap diagnostic.
3. **Weighting estimators** such as IPW, CBPS, and entropy balancing.
4. **Estimator comparison** on the LaLonde data.

A practical note: the comparison functions from the Imbens-Xu tutorial are written for the **ATT**. To stay consistent with that workflow, the estimator comparison section below also focuses on the ATT.

## Setup

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 4.5)

required_packages <- c(
  "tidyverse",
  "estimatr",
  "grf",
  "CBPS",
  "hbal"
)

missing_packages <- required_packages[!required_packages %in% rownames(installed.packages())]
if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

library(tidyverse)
library(estimatr)
library(grf)
library(CBPS)
library(hbal)

# Source the tutorial helper functions
source("https://raw.githubusercontent.com/xuyiqing/lalonde/main/tutorial/functions.R")

# Download the LDW data if it is not already in your working directory
if (!file.exists("lalonde.RData")) {
  download.file(
    url = "https://raw.githubusercontent.com/xuyiqing/lalonde/main/tutorial/data/lalonde.RData",
    destfile = "lalonde.RData",
    mode = "wb"
  )
}

load("lalonde.RData")

# Variables used throughout the lab
Y <- "re78"
treat <- "treat"
covar <- c(
  "age", "education", "black", "hispanic", "married",
  "nodegree", "re74", "re75", "u74", "u75"
)
```

## 1. Meet the data

The loaded file contains several datasets. The two most useful for this lab are:

- `ldw`: the experimental LDW sample.
- `ldw_cps`: the treated LDW units combined with nonexperimental CPS controls.

The experimental sample gives us a useful benchmark, while `ldw_cps` is the observational sample where design problems are much more visible.

```{r data-overview}
tibble(
  sample = c("LDW experimental", "LDW-CPS1"),
  n = c(nrow(ldw), nrow(ldw_cps)),
  treated = c(sum(ldw$treat == 1), sum(ldw_cps$treat == 1)),
  control = c(sum(ldw$treat == 0), sum(ldw_cps$treat == 0))
)
```

```{r look-at-variables}
ldw_cps %>%
  dplyr::select(treat, all_of(covar), all_of(Y)) %>%
  slice(1:6)
```

## 2. A simple benchmark: difference in means

We start with the raw treated-control difference in earnings in 1978.

```{r diff-in-means}
diff(ldw, Y, treat)
diff(ldw_cps, Y, treat)
```

The first estimate comes from the experimental sample; the second comes from the observational sample. The gap between them already tells us that naive comparisons can move a lot once the control group changes.

## 3. Why propensity scores help as a diagnostic

A key problem in observational data is that overlap is hard to inspect in a 10-dimensional covariate space. The propensity score compresses those covariates into a single number,

\[
\hat e(X_i) = \widehat{P}(T_i = 1 \mid X_i),
\]

which makes overlap easier to visualize.

### 3.1 Before the propensity score: raw covariate distributions

Here we compare a few important covariates directly.

```{r raw-covariate-overlap, fig.height=6}
plot_df <- ldw_cps %>%
  select(treat, age, education, re74, re75) %>%
  pivot_longer(cols = c(age, education, re74, re75),
               names_to = "variable", values_to = "value") %>%
  mutate(group = if_else(treat == 1, "Treated", "Control"))

ggplot(plot_df, aes(x = value, fill = group)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  labs(
    title = "Raw covariate distributions in LDW-CPS1",
    x = NULL,
    y = "Density",
    fill = NULL
  ) +
  theme_minimal()
```

This is already useful, but it is still a collection of separate univariate views. It does not solve the multivariate overlap problem.

### 3.2 With the propensity score: one-dimensional overlap plots

```{r assess-overlap, fig.height=4.5}
par(mfrow = c(1, 2))
ldw_ps <- assess_overlap(data = ldw, treat = treat, cov = covar)
ldw_cps_ps <- assess_overlap(data = ldw_cps, treat = treat, cov = covar)
```

The experimental sample should show much better overlap than the observational LDW-CPS1 sample.

**Important:** the propensity score does **not create** overlap where there is no support. What it does is compress the covariates into a balancing score, which makes lack of common support easier to diagnose.

## 4. A manual propensity score weighting example

The tutorial's `ipw()` function estimates ATT weights using a probability forest. Here we reproduce the same basic logic step by step so you can see what the weights are doing.

For the ATT, treated units receive weight 1, and controls receive

\[
\frac{\hat e_i}{1-\hat e_i}.
\]

```{r manual-att-ipw}
ps_model <- probability_forest(
  X = ldw_cps[, covar, drop = FALSE],
  Y = as.factor(ldw_cps[, treat]),
  seed = 1234
)

ldw_cps$ps_hat <- ps_model$predictions[, 2]
ldw_cps$w_att_ipw <- ifelse(ldw_cps$treat == 1, 1, ldw_cps$ps_hat / (1 - ldw_cps$ps_hat))

summary(ldw_cps$ps_hat)
summary(ldw_cps$w_att_ipw)
```

Now use those ATT weights in a weighted regression of the outcome on treatment.

```{r weighted-att-estimate}
ipw_manual <- lm_robust(re78 ~ treat, data = ldw_cps, weights = w_att_ipw, se_type = "stata")
summary(ipw_manual)$coefficients["treat", ]
```

### 4.1 Did weighting improve balance?

A quick way to see this is to compare treated means to control means before and after weighting.

```{r balance-before-after}
weighted_mean <- function(x, w) {
  sum(x * w) / sum(w)
}

balance_tbl <- map_dfr(covar, function(v) {
  x <- ldw_cps[[v]]
  treated <- ldw_cps$treat == 1
  control <- ldw_cps$treat == 0

  tibble(
    variable = v,
    treated_mean = mean(x[treated]),
    control_mean_unweighted = mean(x[control]),
    control_mean_weighted = weighted_mean(x[control], ldw_cps$w_att_ipw[control])
  )
})

balance_tbl
```

The weighted control means should usually move closer to the treated means. That is the design goal of weighting.

## 5. Regression versus outcome modeling using regression

These two methods sound similar but they are doing different things.

### Regression adjustment (`reg`)

This fits one outcome model to the full sample:

\[
Y_i = \alpha + \tau T_i + X_i'\beta + u_i.
\]

The estimated treatment effect is the coefficient on `treat`.

### Outcome modeling using regression (`om.reg`)

This fits an outcome model **only among the controls**:

\[
Y_i = \alpha + X_i'\beta + u_i \qquad \text{for } T_i = 0,
\]

then uses that fitted model to predict the untreated outcomes for the treated units. The ATT is estimated by comparing:

- the observed treated outcomes, and
- the predicted untreated outcomes for those same treated units.

So the difference is:

- `reg` adjusts for covariates in **one pooled regression**.
- `om.reg` uses the controls to **impute the treated units' counterfactual untreated outcome**.

## 6. Estimator comparison on LDW-CPS1

Now we reproduce a simplified version of the comparison figure from the tutorial using only the methods requested for this lab:

- `diff`
- `reg`
- `om.reg`
- `ipw`
- `cbps`
- `ebal`

```{r estimate-selected-methods}
methods_keep <- c("diff", "reg", "om.reg", "ipw", "cbps", "ebal")
labels_keep <- c("Diff-in-Means", "Reg", "OM: Reg", "IPW", "CBPS", "Ebal")

# Experimental benchmark
out_exp <- estimate_all(
  data = ldw,
  Y = Y,
  treat = treat,
  covar = covar,
  methods = methods_keep
)

# Observational sample
out_cps <- estimate_all(
  data = ldw_cps,
  Y = Y,
  treat = treat,
  covar = covar,
  methods = methods_keep
)

out_exp
out_cps
```

```{r plot-comparison, fig.height=5.2}
par(mfrow = c(1, 1))

benchmark_band <- out_exp["diff", c("CI_lower", "CI_upper")]
benchmark_line <- out_exp["diff", "Estimate"]

plot_coef(
  out = out_cps,
  methods = methods_keep,
  labels = labels_keep,
  band = benchmark_band,
  line = benchmark_line,
  ylim = c(-15000, 5500),
  main = "(A) LDW-CPS1"
)
```

## 7. Reading the comparison plot

Use the figure above as follows:

- The **red solid line** is the experimental benchmark ATT estimate.
- The **shaded red band** is the benchmark confidence interval.
- Each black point and interval is one estimator applied to the observational LDW-CPS1 sample.

Questions to ask:

1. Which methods move the estimate closer to the experimental benchmark?
2. Which methods remain far away from the benchmark?
3. Do weighting-based methods look more stable than the naive difference in means?
4. How much do the results depend on overlap and balance?

## 8. Suggested extension

Repeat the same workflow for `ldw_psid` instead of `ldw_cps`.

```{r extension-template, eval=FALSE}
out_psid <- estimate_all(
  data = ldw_psid,
  Y = Y,
  treat = treat,
  covar = covar,
  methods = methods_keep
)

plot_coef(
  out = out_psid,
  methods = methods_keep,
  labels = labels_keep,
  band = benchmark_band,
  line = benchmark_line,
  ylim = c(-15000, 5500),
  main = "(B) LDW-PSID1"
)
```

## 9. Main takeaway

This lab should make three ideas concrete:

1. The propensity score gives a **one-dimensional diagnostic** for overlap.
2. Weighting methods differ in **how they construct weights**, but they all aim at better balance.
3. In observational data, different estimators can behave very differently when overlap is poor.
